---
title: "SIDER 2 Pairwise Similarity"
author: "Daniel Himmelstein"
date: "January 30, 2015"
output:
  html_document:
    theme: cosmo
    highlight: pygments
    includes:
      in_header: ../include/in_header.html
      before_body: include/before_body.html
bibliography: references.bib
csl: ../cite-styles/plos-mod.csl
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(
  fig.width=10,
  fig.path='figures/',
  warning=FALSE,
  message=FALSE,
  fig.retina=2
)
```

```{r, include=FALSE}
# Set up citation management
library(knitcitations)
knitcitations::cleanbib()
knitcitations::cite_options(citation_format = 'pandoc', cite.style = 'authoryear', hyperlink = 'to.bib', super = TRUE)
```

Here we show how to compute the side effect similarity of two drugs. We will be using a similar technique to a previous study which used side effect similarity to uncover new drug-target relationships `r knitcitations::citep('10.1126/science.1158140')`.

### Load packages

We use [*dplyr*](http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html) and [*reshape2*](http://seananderson.ca/2013/10/19/reshape.html) for data manipulation; [*ggplot2*](http://docs.ggplot2.org/current/) and [*ggdendro*](https://github.com/andrie/ggdendro) for plotting; and [*DT*](https://rstudio.github.io/DT/) for displaying html tables using the Javascript [*DataTables* library](http://www.datatables.net/).

```{r, message=FALSE, warning=FALSE}
library(reshape2)
library(dplyr)
library(ggplot2)
library(ggdendro)
library(DT)

options(stringsAsFactors=FALSE)
doMC::registerDoMC(cores = 6)

write.delim <- function(x, file, sep='\t', quote = FALSE, row.names=FALSE, na = '', ...) {
  write.table(x = x, file = file, sep=sep, quote=quote, row.names=row.names, na=na, ...)
}
```


```{r, echo=FALSE}
# Display as a javascript datatable
renderPubChem <- DT::JS(
  "function(data, type, full, meta) {",
  "return type === 'display' ? ",
  "'<a href=\"https://pubchem.ncbi.nlm.nih.gov/compound/' + ",
  "data + '\" target=\"_blank\">' + data + '</a>' : data;}")

DataTable <- function(df, dt.opts = list(), max.rows = 0) {
  if (max.rows != 0) {
    df <- df %>% dplyr::filter(row_number() <= max.rows)
  }
  DT::datatable(data = df, options = dt.opts)
}

PubchemDataTable <- function(df, ...) {
  dt.opts <- list(columnDefs =list(list(targets = which(colnames(df) == 'pubchem_cid') - 1, render = renderPubChem)))
  DataTable(df, dt.opts = dt.opts, ...)
}
```

### Chemical Similarity
We use PubChem's [score matrix](http://pubchem.ncbi.nlm.nih.gov/score_matrix) service. We use the [2D substructure search](https://pubchem.ncbi.nlm.nih.gov/score_matrix/score_matrix-help.html):

> 2D Similarity: Substructure key-based 2D Tanimoto similarity. Scores are in the range [0 .. 100].

Under `ID List`, select *Upload a file with IDs...* and choose `sider_compounds_pubchem.txt`. Select `Id-Id Score` for output format and `gzip`for compression.

```{r}
structure.df <- file.path('..', 'data', 'pubchem_score_matrix_2d.tab.gz') %>%
  read.table(col.names = c('compound_1', 'compound_2', 'substructure')) %>%
  dplyr::mutate(substructure = substructure / 100) %>%
  dplyr::filter(compound_1 != compound_2)

```

### Read data

We begin with the processed data from the [SIDER 2 resource](http://sideeffects.embl.de/) `r knitcitations::citep('10.1038/msb.2009.98')`.

```{r}
sider.df <- file.path('..', 'data', 'sider2-processed.txt') %>% read.delim(na.strings='')
PubchemDataTable(sider.df, max.rows=200)

# Convert into matrixes
se.mat <- xtabs(formula = ~ pubchem_cid + concept_id ,
  data = dplyr::filter(sider.df, type == 'side_effect')) %>% as.matrix()
ind.mat <- xtabs(formula = ~ pubchem_cid + concept_id ,
  data = dplyr::filter(sider.df, type == 'indication')) %>% as.matrix()
```


### Compute concept rarity weights

Rarity weighting was performed as [previousely described](http://www.sciencemag.org/content/suppl/2008/07/10/321.5886.263.DC1/Campillos.SOM.pdf) `r knitcitations::citep('10.1126/science.1158140')`:

> Our weighting scheme consists of two parts, a rareness weight and a correlation weight. The recorded side effects vary greatly in abundance: some, like megaloblastic 4 anaemia, occur in only a few package inserts, while others, like dizziness, appear in most. To be able to account for this, we examined the relation between side-effect frequency and the probability of sharing a drug target within a reference set of 502 drugs with 4857 known human drug–target relations (Fig. S1). We observed an inverse correlation between side-effect frequency and the likelihood of two drugs sharing a protein target. Consequently, the rareness weight for a side effect, $r_i$, is defined as the negative logarithm of the side-effect frequency (Fig. S1D). (This frequency refers to the fraction of package inserts that feature a certain side effect, not the relative occurrence of the side effect in patients.)

```{r, fig.height=3, fig.retina=2}
rarity.se.vec <- -log(colMeans(se.mat))
rarity.ind.vec <- -log(colMeans(ind.mat))

dplyr::bind_rows(
  data.frame(rarity=rarity.se.vec, type='side_effect'),
  data.frame(rarity=rarity.ind.vec, type='indication')) %>%
  ggplot(aes(x = rarity)) + theme_bw() +
  facet_grid(. ~ type) +
  geom_histogram()
```

### Compute concept underrepresentation weights

Underrepresentation weighting was also performed as [previousely described](http://www.sciencemag.org/content/suppl/2008/07/10/321.5886.263.DC1/Campillos.SOM.pdf) `r knitcitations::citep('10.1126/science.1158140')`, except that Ward's method `r knitcitations::citep('10.2307/2282967')` was used for heirarchical clustering:

> Not all side effects are independent of each other; for example, 90% of drugs that cause nausea also cause vomiting. We correct for this redundancy by weighting side effects in a manner analogous to the down-weighting of similar protein sequences within multiple alignments `r knitcitations::citep('10.1016/0022-2836(94)90012-4')` (Fig. S1C). In order to determine the correlation weight, the correlation of side effects was determined by clustering all side effects according to their assigned drugs using a Tanimoto/Jacquard score to compute a distance matrix: The distance between two ~~drugs~~ [side effects] was calculated by dividing the number of drugs that feature both side effects by the number of drugs that have either side effect associated. The Gerstein–Sonnhammer–Chothia algorithm `r knitcitations::citep('10.1016/0022-2836(94)90012-4')` was used to compute weights based on a hierarch[ic]al clustering with the aforementioned distance matrix `r knitcitations::citep('10.1073/pnas.95.25.14863')`.

```{r, fig.height=4, fig.retina=2}
# Uniform wieghting until a better scheme is implemented
underrep.se.vec <- rep(1, ncol(se.mat))
underrep.ind.vec <- rep(1, ncol(ind.mat))

GitHubScript <- function(...) {
  # Source a script from GitHub
  library(RCurl)
  github.url <- file.path('https://raw.githubusercontent.com', ...)
  script <- RCurl::getURL(github.url)
  eval(parse(text = script), envir = .GlobalEnv)
}

CalcUnderrep <- function(mat) {
  # Returns underrepresentation weight for each column
  col.dist <- stats::dist(t(mat), method = 'binary')
  col.clust <- hclust(col.dist, method = 'ward.D2')
  col.dendro <- as.dendrogram(col.clust)
  GitHubScript('antoine-lizee', 'R-GSC', 'master', 'GSC.R')
  gsc.weight <- GSC(col.dendro)
  list(dist = col.dist, clust = col.clust, dendro = col.dendro, weight = gsc.weight)
}

underrep.se <- CalcUnderrep(se.mat)
underrep.ind <- CalcUnderrep(ind.mat)

ggdendro.se <- ggdendro::dendro_data(underrep.se$clust)
ggdendro.ind <- ggdendro::dendro_data(underrep.ind$clust)

dendro.df <- dplyr::bind_rows(
  ggdendro.se %>% ggdendro::segment() %>% dplyr::mutate(type='Side effects'),
  ggdendro.ind %>% ggdendro::segment() %>% dplyr::mutate(type='Indications'))

leaf.df <- dplyr::bind_rows(
  ggdendro.se %>% ggdendro::label() %>% dplyr::left_join(
    y = data.frame(label = names(underrep.se$weight), gsc_weight = underrep.se$weight)) %>% 
    dplyr::mutate(type='Side effects'),
  ggdendro.ind %>% ggdendro::label() %>% dplyr::left_join(
    y = data.frame(label = names(underrep.ind$weight), gsc_weight = underrep.ind$weight)) %>% 
    dplyr::mutate(type='Indications'))

# plot denogrograms
dendro.df %>% ggplot(aes(x=x, y=y)) + theme_bw() +
  facet_grid( ~ type, scales = 'free_x', space = 'free_x') +
  geom_segment(aes(xend=xend, yend=yend), alpha=0.25) +
  geom_point(data = leaf.df, aes(color=gsc_weight), alpha=0.04, size=3) +
  xlab(NULL) + ylab('Distance') +
  scale_color_gradientn(colours=c('#268bd2', '#dc322f'), name='GSC Weight') +
  scale_x_continuous(breaks=NULL, expand = c(0.02, 0)) + scale_y_continuous(breaks=NULL, expand = c(0.04, 0))
```

```{r, fig.height=3, fig.retina=2}
# plot underrepresentation distributions
dplyr::bind_rows(
  data.frame(underrepresentation=underrep.se$weight, type='side_effect'),
  data.frame(underrepresentation=underrep.ind$weight, type='indication')) %>%
  ggplot(aes(x = underrepresentation)) + theme_bw() +
  facet_grid(. ~ type) +
  geom_histogram()
```

### Compute Side Effect/Indication Similarity

Next, we compute weighted [Jaccard index](https://en.wikipedia.org/wiki/Jaccard_index) for compound pairs. This number between 0 and 1 measures the similarity between two side effect/indication profiles. 

```{r}
SiderSimilarity <- function(x1, x2, w) {
  # Compute the similarity between two boolean vectors
  # using a weighted jaccard score.
  sum(w[x1 & x2]) / sum(w[x1 | x2])
}

GetSimilarityDF <- function(mat, w) {
  mat %>%
    proxy::simil(
      method = SiderSimilarity, w = w,
      by_rows = TRUE, upper = FALSE, diag = FALSE) %>%
    as.matrix() %>%
    reshape2::melt(na.rm = TRUE, value.name = 'similarity', varnames = c('compound_1', 'compound_2')) %>%
    dplyr::filter(compound_1 != compound_2) %>% # remove self-similarity (for safety, since `diag = FALSE` in proxy::simil)
    dplyr::distinct(compound_1, compound_2) # proxy::simil returning a non-NA upper half, so duplicate rows exist
}

similarity.df <- dplyr::full_join(
  x = GetSimilarityDF(mat = se.mat, w = rarity.se.vec * underrep.se$weight) %>%
    dplyr::rename(side_effect = similarity),
  y = GetSimilarityDF(mat = ind.mat, w = rarity.ind.vec * underrep.ind$weight) %>%
    dplyr::rename(indication = similarity)) %>%
  dplyr::full_join(y = structure.df) %>%
  dplyr::arrange(compound_1, compound_2)

gz <- file.path('..', 'data', 'similarities.txt.gz') %>% gzfile('w')
similarity.df %>% write.delim(gz); close(gz)
gz <- file.path('..', 'data', 'similarities-complete.txt.gz') %>% gzfile('w')
similarity.df %>% na.omit() %>% write.delim(gz); close(gz)
```

Next, we evaluate the distirubtion of compound-compound similarities. Taking the squareroot of similarities provides a zero-inflated but otherwise normal-looking distribution.

```{r, fig.width=9}
# histogram of similarities
similarity.df %>%
  na.omit() %>%
  reshape2::melt(value.name = 'similarity', variable.name = 'type',
                 id.vars = c('compound_1', 'compound_2')) %>%
  ggplot(aes(similarity)) +
  facet_grid(. ~ type) + 
  geom_histogram(binwidth=0.02)  +
  scale_x_sqrt(breaks = seq(0, 1, 0.2)) + scale_y_sqrt() + theme_bw() +
  xlab('Compound-Compound Similarity')
```

### Does side effect similarity predict indication similarity?

Next, we evaluate whether compound similarity measured by indications is associated with compound similarity measured by side effects.

```{r}
# binned frequency plots
similarity.df %>% na.omit() %>%
  ggplot(aes(side_effect, indication)) + theme_bw() +
    stat_bin2d(aes(alpha=..count.., fill=..count..), bins = 75) + 
    scale_x_sqrt() + scale_y_sqrt() + coord_equal() +
    scale_fill_gradientn(colours=c('#2aa198', '#6c71c4', '#d33682'), trans='log10', name='Compound Pairs') +
    scale_alpha_continuous(range = c(0.3, 1), trans='log10', guide=FALSE) +
    xlab('Side Effect Similarity') + ylab('Indication Similarity')

```

Finally, we fit a linear regression to assess the strength of the association. The association is positive, but has a small effect size. The great significance occurs despite the small effect size since there are close to one million observations.

```{r}
similarity.df %$% lm(sqrt(indication) ~ sqrt(side_effect)) %>% summary()
```

### Does structural similarity predict indication or side effect similarity?

```{r}
similarity.df %>% na.omit() %>%
  reshape2::melt(value.name = 'similarity', measure.vars=c('indication', 'side_effect'), variable.name = 'type') %>%
  ggplot(aes(substructure, similarity)) + theme_bw() +
  facet_grid(. ~ type) + 
    stat_bin2d(aes(alpha=..count.., fill=..count..), bins = 75) + 
    scale_x_sqrt() + scale_y_sqrt() + coord_equal() +
    scale_fill_gradientn(colours=c('#2aa198', '#6c71c4', '#d33682'), trans='log10', name='Compound Pairs') +
    scale_alpha_continuous(range = c(0.3, 1), trans='log10', guide=FALSE) +
    xlab('Substructure Similarity') + ylab('SIDER 2 Drug Label Similarity')
```

```{r}
# Predict indication similarity using substructure similarity
similarity.df %$% lm(sqrt(indication) ~ sqrt(substructure)) %>% summary()
# Predict side effect similarity using substructure similarity
similarity.df %$% lm(sqrt(side_effect) ~ sqrt(substructure)) %>% summary()
# Predict indication similarity using substructure and side effect similarity
similarity.df %$% lm(sqrt(indication) ~ sqrt(substructure) + sqrt(side_effect)) %>% summary()
```

# References
```{r, include=FALSE}
knitcitations::write.bibtex(file = 'references.bib')
```